---
title: "Getting Started with timescapemetrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with timescapemetrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5
)
```

`timescapemetrics` calculates simple, interpretable summaries of binary
time series. For example, consider the case of habitat availability for an imaginary species. 
A time series of habitat availability is represented as a numeric vector
where `1` means habitat is available at a time step and `0` means habitat is
not available.

This vignette introduces the basic workflow:

1. Create or import a binary time series.
2. Calculate one or more temporal metrics.
3. Compare metric values across temporal scenarios.
4. Interpret what the metrics say about habitat persistence, gaps, and
   fragmentation through time.

## Load the package

```{r load-package}
library(timescapemetrics)
```

## Start with one time series

Here is a small example with two periods of habitat availability separated by
unavailable time steps. Each time step in this case represents a year, so the first three years 
have available habitat, then two years of unavailable habitat, followed by three 
more years of available habitat and one year of unavailable habitat.

```{r example-series}
x <- c(1, 1, 1, 0, 0, 1, 1, 0)
x
```

The values can be visualized as a simple timeline.

```{r plot-example-series}
plot(
  seq_along(x), x,
  type = "s",
  ylim = c(-0.1, 1.1),
  xlab = "Time step",
  ylab = "Habitat available",
  yaxt = "n",
  main = "Binary habitat availability through time"
)
axis(2, at = c(0, 1), labels = c("No", "Yes"), las = 1)
```


## Calculate all metrics

Use `ts_metrics()` to calculate the default set of metrics for one series.

```{r all-metrics}
ts_metrics(x)
```

The result is a named numeric vector. Each value summarizes a different aspect
of temporal structure:

| Metric | What it describes in terms of habitat availability |
| --- | --- |
| `total_time` | Number of available time steps. |
| `n_periods` | Number of distinct availability periods. |
| `gap_mean` | Mean length of unavailable gaps. |
| `gap_max` | Longest unavailable gap. |
| `duration_sd` | Variability in the lengths of availability periods. |
| `edge_density` | Frequency of transitions between unavailable and available states. |
| `autocorr` | Lag-1 temporal autocorrelation. |
| `aggregation` | Proportion of neighboring time steps with the same state. |
| `core_time_index` | Proportion of available time that remains after removing edge time steps from each availability period. |
| `n_core_periods` | Number of availability periods long enough to contain core time. |

## Calculate selected metrics

If you only need a few metrics, pass their names to `ts_metrics()`.

```{r selected-metrics}
ts_metrics(
  x,
  metrics = c("total_time", "n_periods", "gap_mean", "edge_density")
)
```

For a single metric, use `ts_metric()`.

```{r single-metric}
ts_metric(x, "n_periods")
```

Metrics can also be called directly as functions.

```{r direct-functions}
ts_total_time(x)
ts_gap_max(x)
ts_core_time_index(x, core_buffer = 1)
```

## Compare temporal scenarios

Metrics can be used to compare time series that differ in total
availability, fragmentation, or gap structure. The examples below all have the
same number of time steps, but they arrange available habitat differently.

```{r scenarios}
scenarios <- list(
  continuous = c(0, 0, 1, 1, 1, 1, 1, 1, 0, 0),
  loss_only = c(0, 0, 0, 1, 1, 1, 1, 0, 0, 0),
  fragmented = c(0, 1, 1, 0, 1, 1, 0, 1, 1, 0),
  irregular = c(1, 0, 1, 1, 0, 0, 1, 0, 1, 1)
)
```

```{r plot-scenarios}
op <- par(mar = c(4, 8, 2, 1))
series_gap <- 1.8
series_y <- rev(seq_along(scenarios)) * series_gap

plot(
  NULL,
  xlim = c(1, length(scenarios[[1]])),
  ylim = c(min(series_y) - 0.5, max(series_y) + 1.5),
  xlab = "Time step",
  ylab = "",
  yaxt = "n",
  main = "Example temporal scenarios"
)

for (i in seq_along(scenarios)) {
  lines(
    seq_along(scenarios[[i]]),
    scenarios[[i]] + series_y[i],
    type = "s",
    lwd = 2
  )
}

axis(
  2,
  at = series_y + 0.5,
  labels = names(scenarios),
  las = 1,
  tick = FALSE
)
par(op)
```

Calculate the same metrics for every scenario and combine the results into a
data frame.

```{r scenario-metrics}
scenario_metrics <- do.call(
  rbind,
  lapply(names(scenarios), function(name) {
    data.frame(
      scenario = name,
      t(ts_metrics(scenarios[[name]])),
      row.names = NULL,
      check.names = FALSE
    )
  })
)

scenario_metrics
```

You can then compare individual metrics across scenarios.

```{r compare-key-metrics}
scenario_metrics[, c(
  "scenario",
  "total_time",
  "n_periods",
  "gap_mean",
  "edge_density",
  "aggregation"
)]
```

In this example, the fragmented scenario has multiple availability periods and
more transitions between `0` and `1`. The continuous scenario has fewer
transitions and higher aggregation because neighboring time steps are more often
in the same state.

## Work with your own data

For now, your input should be a numeric vector containing `0` and `1` values. Missing
values are accepted, but it is usually best to decide how they should be handled
before comparing metrics across time series. For example, if you have a data
frame with one row per time step:

```{r data-frame-example}
habitat <- data.frame(
  time = 1:8,
  available = c(1, 1, 0, 0, 1, 1, 1, 0)
)

ts_metrics(habitat$available)
```

Non-binary values return `NA`, which helps catch accidental use of continuous
or categorical inputs before interpreting the metrics.

```{r invalid-input}
ts_metrics(c(0, 1, 2, 1))
```

## Choosing metrics

Different questions call for different summaries:

- Use `total_time` when you care about the amount of available habitat through
  time.
- Use `n_periods`, `edge_density`, or `aggregation` when you care about temporal
  fragmentation.
- Use `gap_mean` and `gap_max` when interruptions in availability are important.
- Use `core_time_index` and `n_core_periods` when short or edge-dominated
  availability periods may be less useful than sustained availability.

These metrics are intentionally simple. They are designed to make temporal
structure easier to compare across sites, species, simulations, or management
scenarios while keeping the interpretation tied to the original binary time
series.
