---
title: "Getting Started with timescapes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with timescapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`timescapes` summarizes a binary condition through time. A value of `1`
indicates presence and `0` indicates absence. Metric names must match a value
returned by `ts_list_metrics()` exactly.

## A binary vector

```{r vector}
library(timescapes)

x <- c(1, 1, 0, 0, 1, 1, 1, 0)
ts_list_metrics()
ts_metric(x, "total_time")
ts_metric(x, "n_periods")
ts_metric(x, "gap_mean")
```

The two core-time metrics accept a non-negative whole-number buffer:

```{r core}
ts_metric(x, "core_time_index", core_buffer = 1)
ts_metric(x, "n_core_periods", core_buffer = 1)
```

Every observed value must be `0` or `1`. A series containing a missing time
step produces `NA`, because removing or reinterpreting that step could change
the temporal pattern.

```{r missing}
ts_metric(c(1, 1, NA, 0), "n_periods")
```

## A raster time series

A `terra::SpatRaster` can represent the same type of series at many locations.
Layers are treated as ordered time steps.

```{r raster}
library(terra)

r <- rast(nrows = 2, ncols = 2, nlyrs = 4)
values(r) <- matrix(
  c(
    1, 1, 0, 1,
    0, 0, 0, 0,
    1, 0, 1, 0,
    1, 1, 1, 1
  ),
  nrow = ncell(r),
  byrow = TRUE
)

periods <- ts_raster_metric(r, "n_periods")
values(periods)
```

The result is a one-layer raster with the same geometry as the input. Raster
cells containing missing or non-binary values follow the same validation rules
as vector inputs.
