---
title: "Raster workflows with timescapemetrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Raster workflows with timescapemetrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  fig.width = 7,
  fig.height = 4
)
```

```{r load-packages}
library(timescapemetrics)
library(terra)
```

## Overview

Raster layers can represent a time series when each layer stores the state of a
cell at one time step. In `timescapemetrics`, `ts_raster_metric()` applies a
temporal metric to each raster cell, treating the layers as ordered time steps.

This vignette uses a small subset of the Bear River Migratory Bird Refuge in Utah as an example 
dataset included with the package. The raster is binary: `1` indicates water presence and `0` indicates
water absence. The raster was derived from Dynamic World land-cover data (https://www.dynamicworld.app/).
The example also includes a vector layer of the refuge boundary for visualization.

## Load the built-in raster dataset

The example data are stored as packed `terra` objects so they can be included in
the package. Use `terra::rast()` and `terra::vect()` to restore them before
analysis.

```{r load-built-in-data}
data(gsl_bear_river_water)
data(gsl_bear_river_refuge)

water <- rast(gsl_bear_river_water)
refuge <- vect(gsl_bear_river_refuge)

water
refuge
```

Each layer in `water` is one time step. The number of layers is therefore the
length of the time series available for every raster cell.

```{r inspect-raster}
c(
  rows = nrow(water),
  columns = ncol(water),
  cells = ncell(water),
  time_steps = nlyr(water)
)
```

The first layer gives a quick view of the spatial footprint of water presence at
one time step.

```{r plot-input}
plot(water[[1]], main = "Water presence at the first time step")
plot(refuge, add = TRUE)
```

## Calculate a raster metric

Use `ts_raster_metric()` with a metric name from `ts_list_metrics()`. Here,
`n_periods` counts the number of distinct water-presence periods for each cell
through the layer stack.

```{r calculate-n-periods}
water_periods <- ts_raster_metric(water, metric = "n_periods")
water_periods
```

```{r plot-n-periods}
plot(water_periods, main = "Number of water-presence periods")
plot(refuge, add = TRUE)
```

Cells with larger values switched into and out of water presence more often
during the observed time series.

## Summarize metric rasters

Metric rasters can be summarized with standard `terra` tools. For example,
`terra::global()` calculates summary statistics over all cells.

```{r summarize-n-periods}
global(
  water_periods,
  fun = c("mean", "min", "max"),
  na.rm = TRUE
)
```

The raster values can also be extracted as a data frame for plotting or tabular
summaries.

```{r n-periods-frequency}
period_values <- values(water_periods, dataframe = TRUE, na.rm = TRUE)

head(period_values)
table(period_values$`lyr.1`)
```

## Compare several metrics

Different metrics emphasize different aspects of temporal structure. The
examples below calculate the total number of water time steps and the temporal
aggregation of water presence for each cell.

```{r calculate-several-metrics}
water_total_time <- ts_raster_metric(water, metric = "total_time")
water_aggregation <- ts_raster_metric(water, metric = "aggregation")

metric_summary <- rbind(
  total_time = global(water_total_time, "mean", na.rm = TRUE),
  n_periods = global(water_periods, "mean", na.rm = TRUE),
  aggregation = global(water_aggregation, "mean", na.rm = TRUE)
)

metric_summary
```

```{r plot-several-metrics}
plot(
  c(water_total_time, water_periods, water_aggregation),
  main = c("Total water time", "Water periods", "Temporal aggregation")
)
plot(refuge, add = TRUE)
```

`total_time` describes how often water was present. `n_periods` describes how
many separate periods of water presence occurred. `aggregation` describes how
similar adjacent time steps were within each cell's time series.

## Optional class-specific metrics

For categorical rasters, pass `classes` to calculate the same metric separately
for one or more classes. This small in-memory example uses three land-cover
classes across four time steps.

```{r class-specific-example}
classes_raster <- rast(nrows = 2, ncols = 3, nlyrs = 4)
values(classes_raster) <- matrix(
  c(
    1, 1, 2, 2,
    1, 2, 2, 3,
    3, 3, 2, 1,
    2, 2, 1, 1,
    3, 1, 1, 1,
    2, 3, 3, 3
  ),
  nrow = ncell(classes_raster),
  byrow = TRUE
)

class_periods <- ts_raster_metric(
  classes_raster,
  metric = "n_periods",
  classes = c(1, 2, 3)
)

class_periods
values(class_periods)
```

The result has one layer per requested class. Each layer stores the metric value
after converting that class to presence and all other classes to absence.


